\vspace{-15mm}\fontsize{21pt}{10pt}\selectfont\textbf{Wordy Manuscript Title: \\Even Wordier Manuscript Subtitle\thanks{Draft manuscript. Please do not cite without the authors' permission.}}

### 2020-06-18 sections/results.tex
Updates to draft

  We have developed \ac{micone}, a flexible and modular pipeline for 16S amplicon sequencing rRNA data (hereafter mentioned simply as 16S data) analysis, that allows us to infer microbial co-occurence networks.
  It incorporates various popular, publicly available tools as well as custom Python modules and scripts to facilitate inference of co-occurrence networks from 16S data. Using \ac{micone} one can obtain co-occurrence networks by applying to 16S data (or to already processed taxonomic count matrices) any combination of the available tools. The effects of changing any of the intermediate step can be monitored and evaluated in terms of its final network outcome, as well as on any of the intermediate metrics and data outputs.
  The \ac{micone} pipeline workflow is shown in Figure~\ref{fig:figure1}.
  The different steps for going from 16S data to co-occurrence networks can be grouped into four major modules; (i) the denoising and clustering (DC) step, which handles denoising of the raw 16S sequencing data into representative sequences; (ii) the taxonomy assignment (TA) step that assigns taxonomic labels to the representative sequences; (iii) the OTU processing (OP) step that filters and transforms the taxonomy abundance table; and finally (iv) the network inferences (NI) step which infers the microbial co-occurrence network.
  Each process in the pipeline supports alternate tools for performing the same task (see Methods and Figure~\ref{fig:figure1}). A centralized configuration file contains all the specifications for what modules to be used in the pipeline, and can be modified by the user to choose the desired set of tools.
  In what follows, we perform a systematic analysis of each step of the pipeline to estimate how much the final co-occurrence network depends on the possible choices at each step. We also evaluate a large number of tool combinations to determine a set of recommended default options for the pipeline and provide the users with a set of guidelines to facilitate tool selection as appropriate for their data.

@@

  % TODO: Talk about the datasets: real, mock and synthetic (with references)
  In addition to the algorithmic pipeline itself, our analysis involves two types of data: The first type consists of sets of 16S sequencing data for real communities (see Methods). The second are datasets artificially created for the specific goal of helping evaluate computational analysis tools. In particular, in order to objectively compare, to the extent possible, how well each step in \ac{micone} best captures the underlying data, we use both mock data from mockrobiota~\cite{Bokulich2016} as well as, synthetically generated reads from an Illumina read simulator called ART~\cite{Huang2012}.
  The mock datasets selected were labelled mock4, mock12 and mock16 in the mockrobiota database and consist of sequenced 16S data from mock communities of various bacteria.
  These datasets were selected since they contained the expected compositions along with the reference sequences for the organisms in the mock community.
  The synthetic reads were simulated using the three different taxonomy distribution profiles obtained from soil, water and healthy stool microbiome datasets \hl{[ref]}.
  Reference sequences were generated using \ac{ncbi} and the decard tool \cite{Golob2017} for these taxonomy profiles.
  Detailed information on the mock communities and the settings used to generate the synthetic data are provided in the Methods section.

@@

  % TODO: Summarized sections
  In order to analyze the effect of different statistical methods on the inferred co-occurrence networks, we generated co-occurrence networks using all possible combinations of methods and estimated the variability in the networks due to each choice.
  This analysis is performed while keeping the network inference algorithm (NI step) the same throughout the analysis.
  The effects of various steps on the final co-occurrence network is estimated by building a linear model of the edges of the network as a function the various step in the analysis pipeline (see Methods).
  Figure \ref{fig:figure2}B, shows the fraction of total variation among the co-occurrence networks as contributed by the first three steps of the pipeline. In other words, each point corresponds to a different combination of tools, and captures how much the final network is affected by such choice.
  The 16S reference database contributes the most ($\sim25\%$) to variation in the networks. This is also reflected in the fact that the networks can be clearly separated based on the database used (Figure \ref{fig:figure2}B).
  This indicates that the taxonomy assigned to the reference sequences drastically alters the co-occurrence network and this change is much more significant than how the reference sequences themselves are identified (in the DC step).
  We believe that the grouping by taxonomy assignment into clusters (..... in Fig. XXX) is due to the mislabelling of constitutive taxa that are present in high abundance in the community.
  The residual variation can be seen as an artifact that arises due to the effect on the resultant network when multiple steps are changed at the same time.
  Another interesting observation is that the dissimilarity between the networks decreases when the low abundance \ac{otu}s are removed from the network. This is visible in terms of ...
  These results enable us to believe that the most important criterion for accurate comparative analyses of co-occurrence networks is the taxonomy reference database.

@@

  The representative sequences obtained from the \ac{dada2} method in DC step were used for taxonomic assignment using the three reference databases.
  Figure~\ref{fig:figure4}A depicts a flow diagram that shows how the top 50 representative sequences (sorted by abundance) are assigned a Genus according to the three different databases.
  We observe that not only does the assigned Genus composition vary significantly, but the percentage of unassigned representative sequences (gray) also differ.
  Even the most abundant representative sequence is assigned to an "unknown" Genus in two of the three databases.
  A representative sequence might be assigned an "unknown" Genus for one of two reasons: the first is if the taxonomy identifier associated with the sequence in the database did not contain a Genus; the second (more likely) reason is that the database contains multiple sequences that are very similar to the query (representative) sequence and the consensus algorithm (from \ac{qiime2}) is unable to assign one particular Genus at the required confidence.
  After assigning all the representative sequences to taxonomies we perform a pairwise comparison of the similarity between assignments from different databases at every taxonomic level (Figure~\ref{fig:figure4}B).
  The assignments beyond Family level (Family, Genus and Species) are very dissimilar with $<70\%$ similarity between any pair of databases.
  There are no two reference databases that are more similar than the other, with \ac{gg} and SILVA producing only marginally similar assignments compared to \ac{ncbi}.
  This implies that the taxonomy assignments from each reference database are fairly unique and is one of the reasons for a large differences in the resultant co-occurrence networks generated from different taxonomy databases.
   % TODO: Update this
  Supplementary Figure~\ref{fig:figureS4} shows that the top 20 most abundant genera in the three resulting taxonomy composition tables are different.
  For example, the most abundant genus in the \ac{gg} taxonomy table was \textit{Escherichia} whereas in the SILVA taxonomy table it was \textit{Escherichia-Shigella}.
  Although these are minor differences, when comparing a large number of taxonomy composition tables these problems are hard to diagnose.
%   The comparison of all assigned genera instead of the just the top 20 contains the same percentage of matches and mismatches, implying that there does not seem to exist a correlation between abundance and mismatch.
%   This suggests that the most abundant sequences are not necessarily the ones that are consistently matched to the same taxonomies in the different reference databases.

@@

  % TODO: Talk about the difference between correlations and associations
  The six different network inference methods used in this study are \ac{magma}~\hl{in biorxiv [ADD REF]}, \ac{mldm}~\cite{Yang2017}, \ac{spieceasi}~\cite{Kurtz2015}, \ac{sparcc}~\cite{Friedman2012}, Spearman and Pearson.
  These network inference methods into two groups, the first set of methods (Pearson, Spearman, \ac{sparcc}) infer pairwise correlations while the second set infer direct associations (\ac{spieceasi}, \ac{mldm}, \ac{magma}).

>>>>

  We have developed \ac{micone}, a flexible and modular pipeline for 16S amplicon sequencing rRNA data (hereafter mentioned simply as 16S data) analysis, that allows us to infer microbial co-occurrence networks.
  It incorporates various popular, publicly available tools as well as custom Python modules and scripts to facilitate inference of co-occurrence networks from 16S data.
  Using \ac{micone} one can obtain co-occurrence networks by applying to 16S data (or to already processed taxonomic count matrices) any combination of the available tools.
  The effects of changing any of the intermediate step can be monitored and evaluated in terms of its final network outcome, as well as on any of the intermediate metrics and data outputs.
  The \ac{micone} pipeline workflow is shown in Figure~\ref{fig:figure1}.
  The different steps for going from 16S data to co-occurrence networks can be grouped into four major modules; (i) the denoising and clustering (DC) step, which handles denoising of the raw 16S sequencing data into representative sequences; (ii) the taxonomy assignment (TA) step that assigns taxonomic labels to the representative sequences; (iii) the \ac{otu} processing (OP) step that filters and transforms the taxonomy abundance table; and finally (iv) the network inferences (NI) step which infers the microbial co-occurrence network.
  Each process in the pipeline supports alternate tools for performing the same task (see Methods and Figure~\ref{fig:figure1}).
  A centralized configuration file contains all the specifications for what modules to be used in the pipeline, and can be modified by the user to choose the desired set of tools.
  In what follows, we perform a systematic analysis of each step of the pipeline to estimate how much the final co-occurrence network depends on the possible choices at each step.
  We also evaluate a large number of tool combinations to determine a set of recommended default options for the pipeline and provide the users with a set of guidelines to facilitate tool selection as appropriate for their data.

@@

  In addition to the algorithmic pipeline itself, our analysis involves two types of data: The first type consists of sets of 16S sequencing data for real communities consisting of Stool and Oral microbiomes (see Methods).
  The second are datasets artificially created for the specific goal of helping evaluate computational analysis tools.
  In particular, in order to objectively compare, to the extent possible, how well each step in \ac{micone} best captures the underlying data, we use both mock data (labelled mock4, mock12 and mock16) from mockrobiota~\cite{Bokulich2016} as well as, synthetically generated reads from an Illumina read simulator called ART~\cite{Huang2012}.
  These mock datasets were selected since they contained the expected compositions along with the reference sequences for the organisms in the mock community.
  The synthetic reads were simulated using three different taxonomy distribution profiles, namely soil and water microbiomes obtained \ac{emp}~\cite{Thompson2017} and Stool microbiome that is used in our real community analysis~\cite{Kang2017}.
  Reference sequences were generated using \ac{ncbi} and the Decard tool~\cite{Golob2017} for these taxonomy profiles.
  Detailed information on the mock communities and the settings used to generate the synthetic data are provided in the Methods section.

@@

  In order to analyze the effect of different statistical methods on the inferred co-occurrence networks, we generated co-occurrence networks using all possible combinations of methods and estimated the variability in the networks due to each choice.
  This analysis is performed while keeping the network inference algorithm (NI step) the same throughout the analysis.
  The effects of various steps on the final co-occurrence network is estimated by building a linear model of the edges of the network as a function the various step in the analysis pipeline (see Methods).
  Figure \ref{fig:figure2}B, shows the fraction of total variation among the co-occurrence networks as contributed by the first three steps of the pipeline. In other words, each point corresponds to a different combination of tools, and captures how much the final network is affected by such choice.
  The 16S reference database contributes the most ($\sim25\%$) to variation in the networks. This is also reflected in the fact that the networks can be clearly separated based on the database used (Figure \ref{fig:figure2}B).
  This indicates that the taxonomy assigned to the reference sequences drastically alters the co-occurrence network and this change is much more significant than how the reference sequences themselves are identified (in the DC step).
  We believe that the grouping of the networks by taxonomy assignment into clusters (Figure~\ref{fig:figure2}B) is due to the mislabelling of constitutive taxa that are present in high abundance in the community, which drastically alters the nodes and hence the underlying network topology.
  The residual variation (Figure \ref{fig:figure2}A) can be seen as an artifact that arises when multiple steps are changed at the same time.
  Another interesting observation is that the dissimilarity between the networks decreases when the low abundance \ac{otu}s are removed from the network. This result is elaborated in detail in the denoising and clustering section.
  These results enable us to believe that the most important criterion for accurate comparative analyses of co-occurrence networks is the taxonomy reference database.

@@

  The representative sequences obtained from the \ac{dada2} method in DC step were used for taxonomic assignment using the three reference databases.
  Figure~\ref{fig:figure4}A depicts a flow diagram that shows how the top 50 representative sequences (sorted by abundance) are assigned a Genus according to the three different databases.
  We observe that not only does the assigned Genus composition vary significantly, but the percentage of unassigned representative sequences (gray) also differ.
  Even the most abundant representative sequence is assigned to an "unknown" Genus in two of the three databases.
  A representative sequence might be assigned an "unknown" Genus for one of two reasons: the first is if the taxonomy identifier associated with the sequence in the database did not contain a Genus; the second (more likely) reason is that the database contains multiple sequences that are very similar to the query (representative) sequence and the consensus algorithm (from \ac{qiime2}) is unable to assign one particular Genus at the required confidence.
  After assigning all the representative sequences to taxonomies we perform a pairwise comparison of the similarity between assignments from different databases at every taxonomic level (Figure~\ref{fig:figure4}B).
  The assignments beyond Family level (Family, Genus and Species) are very dissimilar with $<70\%$ similarity between any pair of databases.
  There are no two reference databases that are more similar than the other, with \ac{gg} and SILVA producing only marginally similar assignments compared to \ac{ncbi}.
  This implies that the taxonomy assignments from each reference database are fairly unique and is one of the reasons for a large differences in the resultant co-occurrence networks generated from different taxonomy databases.

  Supplementary Figure~\ref{fig:figureS4} shows that the top 20 most abundant genera in the three resulting taxonomy composition tables are different.
  For example, the most abundant genus in the \ac{gg} taxonomy table was \textit{Escherichia} whereas in the SILVA taxonomy table it was \textit{Escherichia-Shigella}.
  Although these are minor differences, when comparing a large number of taxonomy composition tables these problems are hard to diagnose.
%   The comparison of all assigned genera instead of the just the top 20 contains the same percentage of matches and mismatches, implying that there does not seem to exist a correlation between abundance and mismatch.
%   This suggests that the most abundant sequences are not necessarily the ones that are consistently matched to the same taxonomies in the different reference databases.

@@

  % TODO: Talk about the difference between correlations and associations
  The six different network inference methods used in this study are \ac{magma}~\cite{Cougoul2019}, \ac{mldm}~\cite{Yang2017}, \ac{spieceasi}~\cite{Kurtz2015}, \ac{sparcc}~\cite{Friedman2012}, Spearman and Pearson.
  These network inference methods into two groups, the first set of methods (Pearson, Spearman, \ac{sparcc}) infer pairwise correlations while the second set infer direct associations (\ac{spieceasi}, \ac{mldm}, \ac{magma}).
  Pairwise correlation methods involve calculating the correlation coefficient between every pair of \ac{OTU}/\ac{esv}s leading to the detection of spurious indirect connections.
  On the other hand, direct association methods use conditional independence to avoid the detection of correlated but indirectly connected \ac{otu}s.
